---
title: " Example SMILE model Analysis"
author: "Dan McGuire"
date: "`r Sys.Date()`"
output: rmarkdown::github_document
vignette: >
  %\VignetteIndexEntry{Selectorgadget}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r,echo=F} 

#library(data.table)
##library(INLA)
#library(msr)
#library(Matrix)
#library(parallel)
#library(sf)
#library(ggplot2)
#library(broom)
##library(inlabru)
#library(sandwich)
## output: rmarkdown::html_vignette
library(knitr)
opts_chunk$set(comment = "#>", collapse = TRUE)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$out.lines)) {
    x = knitr:::split_lines(x)
    if (length(x) > n) {
      # truncate the output
      h = c(head(x, n), '....\n','....\n')
      t = c(tail(x, n))
      x = c(h, t)
    }
    x = paste(x, collapse = '\n') # paste first n lines together
  }
  hook_output(x, options)
})

#opts_chunk$set(eval=FALSE)
#Rscript -e "rmarkdown::render('mamba_example/example_mamba_analysis.Rmd')"

```
# SMILE 
A **S**patial-**M****I**xed **L**inear **E**ffects model for estimating the contributions of genetic heritability, family-, and community-level environment to phenotype variation.


## Installation

    library(devtools)
    install_github("dan11mcguire/smile")


##  Detailed example 
1. Simulate genetic, spatial community-level, and family-level environment variance components.
2. Then, try to recapture the simulated variances with the estimated SMILE model.

##### Part 1.
```{r}
library(smile)
library(data.table)
library(Matrix)
library(parallel)
library(sf)
library(ggplot2)
library(broom)
library(spdep)
library(sandwich)


data(lgeo_opt)
lgeo_opt[1:20]
seed<-808
n_quad=10*10**3
```

Calculate spatial neighborhood matrix
```{r}

sampleloc<-as.character(sample(x=lgeo_opt[,key], 
			       size=n_quad, 
			       prob=lgeo_opt[,prop],replace=TRUE))

ngeo_opt<-lgeo_opt[key %in% sampleloc,head(.SD,1),by=key]
ngeo_opt<-as.data.table(ngeo_opt)[,rn:=1:.N]
wsp<-poly2nb(as(st_as_sf(ngeo_opt), "Spatial"),snap=0.01)
whch<-sapply(1:length(wsp), function(x) ifelse(wsp[[x]][1]!=0,x,-9))
fillin<-which(whch==-9)
whch<-which(whch!=-9)
ivec<-rep(whch, 
	  times=sapply(wsp[whch], length))
jvec<-unlist(lapply(1:length(whch), 
		    function(x){
		      wsp[[whch[x]]] 
		    }))

wspm<-sparseMatrix(i=ivec, j=jvec)
isSymmetric(wspm)
wmatd<-fillin_nearest_neighb(as.data.table(ngeo_opt),wspm)$wnew


rs<-rowSums(wmatd)
print(mean(1/rs))
wmatplus<-as(do.call(rbind, 
    		 lapply(1:nrow(wmatd), function(i) wmatd[i,]/rs[i])), "dgTMatrix")

Mpm5<-as(sparseMatrix(i=1:length(rs), 
		      j=1:length(rs),
		      x=1/rs^-.5),
	 "dgTMatrix")

Mpp5<-as(sparseMatrix(i=1:length(rs), 
		      j=1:length(rs),
		      x=1/rs^.5),
	 "dgTMatrix")

diagI<-as(sparseMatrix(i=1:length(rs), 
		      j=1:length(rs),
		      x=1),
	 "dgTMatrix")
N<-Mpm5%*% wmatplus %*% Mpp5
      

### Zs:  loading matrix mapping individuals to their location-specific spatial random effects  ###
ds<-data.table(kid=rep(sampleloc,each=4))
ds<-merge(ds,ngeo_opt[,.(gid,kid=key,rn)],by=c("kid"),sort=F)    
ivec<-lapply(1:ds[,max(rn)], function(x) ds[rn==x,which=TRUE]) 
jvec<-lapply(1:length(ivec), function(x) rep(x, length(ivec[[x]])))  
Zs<-sparseMatrix(i=unlist(ivec), j=unlist(jvec), x=1,giveCsparse=F)

```

Simulate (S) spatial random effects with CAR covariance structure, and simulate (G) genetic ,(P) parental-, and (C) child- level environment random effects.  

Let G=0.4, P=0.05, C=0.2, S=0.03
```{r}

S<-0.03
rho<-0.9
sigma_inv<-(Mpm5%*%(diagI - rho*N)%*%Mpm5) ### CAR specification of covariance matrix
sigma_s<-chol2inv(chol(sigma_inv))
sets2<-S/(gowfac(Z=Zs,G=sigma_s) / nrow(Zs))
all.equal(t(chol(sigma_inv)) %*% (chol(sigma_inv)), sigma_inv)
summary(apply(sigma_inv, 2, function(x) sum(x!=0)))
summary(apply(sigma_s, 2, function(x) sum(x!=0)))   
chsigma_s<-t(chol(sigma_s))

ustar<-rnorm(nrow(wmatd), sd=1)*sqrt(sets2)
var(ustar)
ustar<-chsigma_s %*% ustar
ustar<-as.vector(Zs %*% ustar)
ds[["u_s"]]<-ustar

G=0.4
Fam=0
P=0.05
C=0.2
E<- 1 - G - Fam - P - C - S

n_trio<-0
mu<-0
print("simulating data.")


ss<-smile::sim_fam_nonspatial(
                        n_quad=n_quad
                       ,Ag=G
      	               ,C_fam=Fam
      	               ,C_par=P
                       ,C_sib=C
      	               ,S=S
                       ,mu=0
                          )


ss$d<-cbind(ss$d, ds)
ss$d[,y:=u_a + u_sib + u_par + u_fam + eps + fixed + u_s]
ss$d[1:8]

ss$d[,lapply(.SD, var),.SDcols=grep("u_|eps",names(ss$d),value=T)]


ss$d[,lapply(.SD, var),.SDcols=c(grep("^[yu]", names(ss$d),value=T))] ## Actual variance of the outcome and the random effects
mlists<-list(Ltlist=ss$Ltlist, Zlist=ss$Zlist,Glist=ss$Glist)
ss$Zlist<-NULL
ss$Glist<-NULL
ss$Ltlist<-NULL


mlists$Glist$Ga[1:12,1:12]
mlists$Zlist$Zc_par[1:12,1:9]
mlists$Zlist$Zc_sib[1:12,1:9]
mlists$Zlist$Zc_fam[1:12,1:3]


vc_struct<-"sgps"
vcl<-strsplit(vc_struct,"")[[1]]
tmplt<-paste0("fam_chol_car_gmrf_",vc_struct,"_us")
#tmplt<-paste0("fam_chol_sar_gmrf_",vc_struct,"_us")
if(vcl[1]!="s"){
  tmplt<-gsub("car","matern",tmplt)
}


K<-0.05
ss$d[,yb:=as.numeric(y > qnorm(1-K))]

 
binaryind<-ifelse(K==1,0,1) 
khat<-ss$d[,mean(yb)]
z<-dnorm(qnorm(1-khat))
yvec<-ss$d[,yb]

```
 
##### Part 2.
Fit the smile model using the `smile()` function like below.

```{r,out.lines=20}
model_output<-
     smile(
       X=matrix(rep(1,nrow(ss$d),ncol=1))
      ,y=yvec
      ,Zs=Zs
      ,wmatd=wmatd
      ,vc_structure="gpcs"
      ,spatial_structure="car"
      ,mlists=mlists
        )
```

`vc_a_lia`, `vc_c_par_lia`, `vc_c_sib_lia`, `gvc_s_lia`, represent the G, P, C, and S variance compoenents respectively.
The model estimates are close to the true simulated variances.
```{r}
model_output$vc_rpt
```
The full model report from TMB with all fitted parameters are found in the `tmb_rpt` object.
```{r}
model_output$tmb_rpt
```

The `indlevel_dt` object is a data-table showing all fitted effects and the response variable.
```{r}
model_output$indlevel_dt
```
















